---
title: 'Final Project'
author: "Patkhedkar Ninad"
date: "2020-11-13"
output:
  pdf_document: default
  html_document: default
---

## How to import and clean my data?

I have 3 datasets. Lets check and clean up dataset one by one

### Dataset file - jeee16t03.csv

First We will simplify the labels as labels contains spaces and some names are too long. We am replacing spaces with underscores and changing to lower case. We will also rename 2 columns
- state_and_type_of_government change to "state"
- population_2016_thousands to "population_k"

```{r, message=FALSE, warning=FALSE, echo = TRUE}
library(knitr)
library(dplyr)
library(janitor)

options("width"=200)
options(scipen=999)  # turn-off scientific notation like 1e+48
setwd("/cloud/project/completed/final_project")

df_jee03 <- read.csv("jeee16t03.csv")
df_jee03 <- df_jee03 %>% clean_names() %>%
              rename(state = state_and_type_of_government) %>%
              rename(population_k = population_2016_thousands)
```

We will also just focusing only on records from each state. There are some records at other levels like local county govt, muncipalty and no population is provided for such records. So We will dropping all such records keeping only State govt level records.

```{r, message=FALSE, warning=FALSE, echo = TRUE}
df_jee03 <- df_jee03 %>% filter(population_k != "-")
df_jee03 <- df_jee03[-1,]
head(df_jee03[,1:5],10)
```

First record in dataset is total of all taxes. We will drop that record.
All state entries are followed by 2 letter abbreviation like Virginia (VA). We will trim these 2 letters abbreviations. Will also trim and leading and trailing spaces.
```{r, message=FALSE, warning=FALSE, echo = TRUE}
df_jee03$state <- sub("\\(.*", "",df_jee03$state )
trim <- function (x) gsub("^\\s+|\\s+$", "", x)
df_jee03$state <- trim(df_jee03$state)
head(df_jee03[,1:5],10)
```

### Dataset file - jeee16t08.csv

Again will simplify columns names by changing to lower case and replacing spaces with underscores. We will change the column name "population_2016" to "popuation" as all data is of year 2016.

I will also drop firest record as it talks about Total of all states. I am focused on individual state data.
```{r, message=FALSE, warning=FALSE, echo = TRUE}
df_jee08 <- read.csv("jeee16t08.csv")
df_jee08 <-df_jee08 %>% clean_names() %>%
               rename(population = population_2016)

df_jee08 <-filter(df_jee08, state != "Total")
head(df_jee08[,1:5])
```

### Dataset file - jeee16t11.csv
Again will simplify columns names by changing to lower case and replacing spaces with underscores.
We will drop first record as its for Total of all states. We will concentrate on statewise data.
```{r, message=FALSE, warning=FALSE, echo = TRUE}
df_jee11 <- read.csv("jeee16t11.csv")
df_jee11 <- df_jee11 %>% clean_names() %>% filter(state != "Total")
head(df_jee11[,1:5]) 
```

## What does the final data set look like?
We will consolidate all datasets by joining together on "state" field.
```{r, message=FALSE, warning=FALSE, echo = TRUE}
df_consolidated <- inner_join(df_jee08, df_jee03, by = "state")
df_consolidated <- inner_join(df_consolidated, df_jee11, by = "state")
```
We got 2 population fields in consolidated dataset.

population_k  - represents population of state in thousands ...basically round figure(k - stands for 1000)
population -  represents actual count of population

We will keep field which represents population in thousands as its easy to follow for analysis. We will drop other population field.

Then we will check the details of all fields in our dataframe.

```{r, message=FALSE, warning=FALSE, echo = TRUE}
df_consolidated <- df_consolidated %>% select(-c(population))
str(df_consolidated)
```

## Questions for future steps.
Considering the questions we want to find answer for, its required to identify correct variables. Currently there are 41 variables after joining the datasets.

I have identfied below variables which we would use. However based on how analysis goes, we may need to add or drop some variables.

- state
- population_k
- total_direct_expenditure
- police_protection_amount
- total_justice_system_amount
- total_justice_system_pc
- tjs_total_employees
- tjs_full_time_equivalent
- tjs_average_earnings

## What information is not self-evident?
There is no crime rate related data in datasets. It can be assumed that Police protection functions cost more in states having large metro areas with high crime rate.But its not clear if civil services expense are also high in such states. We will try to establish correlation between spending on police protection and civil services.

## What are different ways you could look at this data? 
I plan to perform linear regression and correlation analysis to find some variables which may have impact expenses.We will also explore clustering based on police costs. 

## How do you plan to slice and dice the data? 
Yes. Dataset is already created by joining two datsets. We may need to further derive employment related variable by combining full time and part time data.

## How could you summarize your data to answer key questions? 
```{r, message=FALSE, warning=FALSE, echo = TRUE}
library(skimr)
#skim(df_consolidated)
summary(df_consolidated)
```
For some reason, skim output is not showing up in R Markdown. However summary data is good enough.

## What types of plots and tables will help you to illustrate the findings to your questions?
Scatter plots, Histograms and Cluster plots will help with findings on questions.

## Do you plan on incorporating any machine learning techniques to answer your research questions? Explain. 
Yes, I plan to use k clustering for classification of states on some metrics like salaries etc.
